---
title: "FY12-24YTD ICE detention analysis of date fields"
author: "UWCHR"
date: "2024-10-09"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

This notebook analyzes characteristics of datetime columns of ICE detention records from October 1, 2011 to January 4, 2024. As discussed in the [dataset overview](index.html), each row in the detention records dataset represents a "placement" of an individual detained person at a specific detention facility, demarcated by start and end booking dates (`detention_book_in_date_and_time`, `detention_book_in_date_time`). One or more successive placement records for an individual constitute a detention "stay", covering the entire period of their detention. Again, these are demarcated with start and end dates (`stay_book_in_date_time`, `stay_book_out_date_time`).

```{r import_data, message=FALSE, warning=FALSE, include=TRUE}
options(scipen = 1000000)

library(pacman)
p_load(tidyverse, arrow, lubridate, zoo, digest, ggplot2, plotly, gghighlight, knitr, kableExtra)

df <- data.table::fread(file = here::here('export', 'output', 'ice_detentions_fy12-24ytd.csv.gz'),
                        sep = '|')

data.table::setDF(df)

detloc_aor <- df %>% 
  distinct(detention_facility_code, area_of_responsibility) %>% 
  arrange(detention_facility_code, area_of_responsibility)

# Need to look more closely at which records are retained here
# Assume it is first placement per stay
# unique_stays <- df %>% 
#   distinct(stayid, .keep_all = TRUE)

```

```{r duplicate_placements}

# Some cases have duplicate placement records: two book-ins at same time. For most (not all!) of these, one of the duplicate placement records has placement length equal to zero.
# These may represent data entry corrections or additions; details in detained person characteristics columns appear to differ in many cases

# Flag duplicate placement book-ins within each stay
df <- df %>% 
  group_by(stayid) %>% 
  mutate(isdupe = duplicated(detention_book_in_date_and_time) | duplicated(detention_book_in_date_and_time, fromLast = TRUE)) %>% 
  ungroup()

alldupes <- df %>% filter(isdupe == TRUE)

dupe_recids <- alldupes$recid

# Check if all zero length placements are duplicates (probably not)

zero_placement <- df %>% 
  filter(placement_length_min == 0)

zero_recids <- zero_placement$recid



```

# Date range

The following table offers an overview of placement and stay book-in and book-out fields. Note this dataset should include all detention records for all people who experienced ICE detention during October 1, 2011 through January 4, 2024; accordingly, this includes people with placements and stays beginning prior to October 1, 2011.

We have interpreted records with null placement and stay book-out values as representing ongoing detentions at the time of generation of the dataset. 

It is logical to observe more null stay book-out values than null placement book-out dates, as each "bloc" of successive placement records associated with an ongoing stay will be present multiple null stay book-out values.

We should not expect to see any records with null placement book-out and non-null stay book-out: `r stopifnot(nrow(df[is.na(df$detention_book_out_date_time) & !is.na(df$stay_book_out_date_time),]) == 0)`.

However, we do see records with no null placement book-out values, but null stay book-out values, indicating that the stay appears to be ongoing but no ongoing placement appears in the dataset.

```{r date_range}

date_cols <- c("stay_book_in_date_time",
               "detention_book_in_date_and_time",
               "detention_book_out_date_time",
               "stay_book_out_date_time",
               "detention_release_reason",
               "stay_release_reason")

skimr::skim(df[date_cols])

skimr::skim(unique_stays[date_cols])

```

```{r test}

test <- unique_stays %>% 
  filter(is.na(stay_book_out_date_time),
         !is.na(detention_book_out_date_time))

test_ids <- test$stayid

test2 <- df %>% 
  filter(stayid %in% test_ids)

```

# Placement length

`r sum(df$placement_length == 0, na.rm = TRUE)` completed placements have a `placement_length` of zero; `r sum(df$placement_length < 0, na.rm = TRUE)` have negative `placement_length` values. 

```{r placement_length}

dat <- df %>% 
  mutate(placement_length_bin = cut(placement_length,
                           breaks = c(-Inf, 0, 3, 30, 90, 360, Inf),
                           right = TRUE))
p1 <- dat %>%
  count(placement_length_bin) %>% 
  ggplot(aes(x = placement_length_bin, y = n)) +
  geom_col()

p1

```

# Stay length

`r sum(unique_stays$stay_length == 0, na.rm = TRUE)` completed stays have a `stay_length` of zero; `r sum(unique_stays$stay_length < 0, na.rm = TRUE)` have negative `stay_length` values. 

```{r stay_length}

dat <- unique_stays %>%
  mutate(stay_length_bin = cut(placement_length,
                           breaks = c(-Inf, 0, 3, 30, 90, 360, Inf),
                           right = TRUE))

p1 <- dat %>%
  count(stay_length_bin) %>% 
  ggplot(aes(x = stay_length_bin, y = n)) +
  geom_col()

p1

```

# Successive placements

Examining characteristics of successive placements, using FY2020 as test case, as this calculation is memory-intensive for full dataset. Some detention placement book-ins have exact same time as previous placement book-out. Others may have book-in prior to previous book-out, or with long gaps between placements.

```{r placement_lag}

test_df <- df %>% 
  filter(detention_book_in_date_and_time >= '2019-10-01',
         detention_book_in_date_and_time <= '2020-09-30')

test_df <- test_df %>%
  group_by(stayid) %>%
  mutate(bookout_lag = lag(detention_book_out_date_time, n=1, order_by=stayid),
         bookin_diff = difftime(detention_book_in_date_and_time, bookout_lag, unit="days")) %>%
  ungroup()

dat <- as.numeric(unlist(test_df[!is.na(test_df$bookin_diff), 'bookin_diff']))

hist(dat, right = T, breaks=c(-100, -1, 0, 1, 250))

```

Examining stay blocs with odd placement timings (instant transfers, impossible placement lengths, long gaps between placements).

Many (but not all) records involving instant transfers (book-out time exactly equal to subsequent book-in time) appear to involve cases with `detention_release_reason` values of "Processing Disposition Changed Locally"; this would make sense if these placements are "on paper only" or simply administrative changes. These cases are not uncommon (`r length(ex_zero_ids)` individual stays during FY2020), so this is a good argument for using a headcount methodology which is not inclusive of book-out dates in order to not double-count these placements (see below).

Negative placement gaps mean that the dataset records an individual as being in two places at one time, e.g. one placement is not completed but subsequent has already started. These are not common (`r length(ex_neg_ids)` individual stays during FY2020).

```{r}

sample_cols <- c("stay_book_in_date_time",
                 "detention_book_in_date_and_time", # Note irregular field name format
                 "detention_book_out_date_time",
                 "stay_book_out_date_time",
                 "citizenship_country",
                 "gender",
                 "birth_year",
                 "detention_facility",
                 "area_of_responsibility",
                 "detention_release_reason",
                 "stay_release_reason",
                 "anonymized_identifier",
                 "stay_length",
                 "placement_length",
                 "bookout_lag",
                 "bookin_diff")

ex_zero <- test_df[test_df$bookin_diff == 0,] %>%
  filter(!is.na(stay_book_in_date_time))

ex_zero_ids <- unique(ex_zero$anonymized_identifier)

df_ex_zero <- test_df %>%
  filter(anonymized_identifier %in% ex_zero_ids) %>% 
  dplyr::select(all_of(sample_cols))


###

ex_neg <- test_df[test_df$bookin_diff < 0,] %>%
  filter(!is.na(stay_book_in_date_time))

ex_neg_ids <- unique(ex_neg$anonymized_identifier)

df_ex_neg <- test_df %>%
  filter(anonymized_identifier %in% ex_neg_ids) %>% 
  dplyr::select(all_of(sample_cols))

###

ex_gap <- test_df[test_df$bookin_diff > 3,] %>%
  filter(!is.na(stay_book_in_date_time))

ex_gap_ids <- unique(ex_gap$anonymized_identifier)

df_ex_gap <- test_df %>%
  filter(anonymized_identifier %in% ex_gap_ids) %>% 
  dplyr::select(all_of(sample_cols))

```

Comparing outcomes of headcount based on inclusive versus exclusive indexing of placement book-out date-time. Exclusive indexing is more conservative; inclusive indexing may over-count if successive placements have exact same book-out and book-in values.

```{r daily_headcount, message=FALSE, warning=FALSE, include=TRUE}

# Date exclusive indexing
headcount_exc <- read_delim(here::here('export', 'output', 'headcount_fy12-24ytd.csv.gz'), delim='|')

# Date inclusive indexing (Note this file not included in GitHub repo)
headcount_inc <- read_delim(here::here('headcount', 'frozen', 'headcount_fy12-24ytd.csv.gz'), delim='|')

# Exclusive of book-out date-time
headcount_exc <- headcount_exc %>% 
  mutate(date = as.Date(date),
         v = "exclusive")

# Inclusive of book-out date-time
headcount_inc <- headcount_inc %>% 
  mutate(date = as.Date(date),
         v = "inclusive")

hc_compare <- rbind(headcount_exc, headcount_inc)

p1 <- hc_compare %>% 
  group_by(date, v) %>% 
  summarize(n = sum(n)) %>% 
  ggplot(aes(x = date, y = n, color = v)) +
  geom_line()

p1

dat <- hc_compare %>% 
  group_by(date, v) %>% 
  summarize(n = sum(n)) %>% 
  pivot_wider(id_cols = date, names_from = v, values_from = n) %>% 
  mutate(diff = exclusive - inclusive)

p2 <- dat %>% 
  ggplot(aes(x = date, y = diff)) +
  geom_line()

p2

```


```{r stay_headcounter, message=FALSE, warning=FALSE, include=TRUE}

# Transform data to "fill in" missing `detention_book_out_date_time` values
# in order to account for ongong detention stays at time of release of data
max_date <- max(df$stay_book_out_date_time, na.rm=TRUE)

unique_stays <- unique_stays %>%
  mutate(stay_book_out_date_time_min = 
           case_when(is.na(stay_book_out_date_time) ~ max_date,
                     TRUE ~ stay_book_out_date_time))

# Define timeline for calculation of daily detained population
timeline_start <- min(unique_stays$stay_book_in_date_time, na.rm=TRUE)
timeline_end <- max(unique_stays$stay_book_out_date_time, na.rm=TRUE)
timeline <- seq(timeline_start, timeline_end, by='day')

# Function counts all in-range detention placement records in dataset `df` for a
# given `date` by a given grouping variable `var`
stay_headcounter <- function(date, df, group_vars) {
  
  in_range <- df[df$stay_book_in_date_time <= date & df$stay_book_out_date_time_min >= date,]
  
  in_range %>% 
    group_by(across(all_of(group_vars))) %>% 
    summarize(n = n()) %>% 
    complete(fill = list(n = 0)) %>% 
    mutate(date=date)
  
  }


# Apply function to timeline
stay_headcount <- lapply(timeline, stay_headcounter, df=unique_stays, group_vars=c('detention_facility_code'))

# Transform output into data frame
stay_headcount_data <- map_dfr(stay_headcount, bind_rows)

write_delim(stay_headcount_data, here::here('analyze', 'frozen', 'stay_headcount_fy12-24ytd.csv.gz'), delim='|')

```

```{r compare_stay_placement_headcounts}

stay_headcount_data <- read_delim(here::here('analyze', 'frozen', 'stay_headcount_fy12-24ytd.csv.gz'), delim='|')


# Plot headcount
p1 <- stay_headcount_data %>%
  group_by(date) %>% 
  summarize(n = sum(n)) %>% 
  filter(date >= "2010-10-01") %>%
  ggplot(aes(x = date, y = n) ) +
  geom_line() +
  labs(title = "Daily detained population")

p1

stay_headcount_daily <- stay_headcount_data %>%
  group_by(date) %>% 
  summarize(n = sum(n)) %>% 
  mutate(v = "stays",
         date = as.Date(date))

headcount_daily <- headcount_exc %>%
  group_by(date) %>% 
  summarize(n = sum(n)) %>% 
  mutate(v = "placements",
         date = as.Date(date))

dat <- rbind(stay_headcount_daily, headcount_daily)

p2 <- dat %>% 
  ggplot(aes(x = date, y = n, color = v)) +
  geom_line()

p2

dat_wide <- dat %>% 
  pivot_wider(id_cols = date, names_from = v, values_from = n) %>% 
  mutate(diff = stays - placements)

p3 <- dat_wide %>% 
  ggplot(aes(x = date, y = diff)) +
  geom_line()

p3

```


```{r headcounter, message=FALSE, warning=FALSE, include=TRUE}

# Transform data to "fill in" missing `detention_book_out_date_time` values
# in order to account for ongong detention stays at time of release of data
max_date <- max(df$stay_book_out_date_time, na.rm=TRUE)

df <- df %>%
  mutate(detention_book_out_date_time_min = 
           case_when(is.na(detention_book_out_date_time) ~ max_date,
                     TRUE ~ detention_book_out_date_time))

# Define timeline for calculation of daily detained population
timeline_start <- min(df$stay_book_in_date_time, na.rm=TRUE)
timeline_end <- max(df$stay_book_out_date_time, na.rm=TRUE)
timeline <- seq(timeline_start, timeline_end, by='day')

# Function counts all in-range detention placement records in dataset `df` for a
# given `date` by a given grouping variable `var`
headcounter <- function(date, df, group_vars) {
  
  in_range <- df[df$detention_book_in_date_and_time <= date & df$detention_book_out_date_time_min >= date,]
  
  in_range %>% 
    group_by(across(all_of(group_vars))) %>% 
    summarize(n = n()) %>% 
    complete(fill = list(n = 0)) %>% 
    mutate(date=date)
  
  }

# # Generate limited sample dataset
# test_df_df <- df %>% 
#   filter(citizenship_country == "ANGOLA")
# 
# # Apply function to timeline
# example_headcount <- lapply(timeline, headcounter, df=test_df_df, group_vars=c('gender'))
# 
# # Transform output into data frame
# example_headcount_data <- map_dfr(example_headcount, bind_rows)
# 
# # Plot headcount
# p1 <- example_headcount_data %>% 
#   filter(date >= "2010-10-01") %>% 
#   ggplot(aes(x = date, y = n, fill = gender) ) +
#   geom_area() +
#   labs(title = "Daily detained population by `gender`",
#        subtitle = "`citizenship_country` == 'ANGOLA'")
# 
# p1

```
