---
title: "FY12-24YTD ICE detention analysis of date fields"
author: "UWCHR"
date: "2024-10-09"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

This notebook analyzes characteristics of datetime columns of ICE detention records from October 1, 2011 to January 4, 2024. As discussed in the [dataset overview](index.html), each row in the detention records dataset represents a "placement" of an individual detained person at a specific detention facility, demarcated by start and end booking datetime values (`detention_book_in_date_and_time`, `detention_book_in_date_time`). One or more successive placement records for an individual constitute a detention "stay", covering the entire period of their detention. Again, these are demarcated with start and end booking datetime values (`stay_book_in_date_time`, `stay_book_out_date_time`).

```{r import_data, message=FALSE, warning=FALSE, include=TRUE}
options(scipen = 1000000)

library(pacman)
p_load(tidyverse, arrow, lubridate, zoo, digest, ggplot2, plotly, gghighlight, knitr, kableExtra)

df <- data.table::fread(file = here::here('export', 'output', 'ice_detentions_fy12-24ytd.csv.gz'),
                        sep = '|')

data.table::setDF(df)

detloc_aor <- df %>% 
  distinct(detention_facility_code, area_of_responsibility) %>% 
  arrange(detention_facility_code, area_of_responsibility)

```

# Duplicate placements

A small number of stays include records with duplicate `detention_book_in_date_and_time` values; a smaller number of stays include duplicate `detention_book_out_date_time` values. In many cases, one of these duplicate placements has `placement_length` equal to zero. These may represent data entry corrections or additions; details in detained person characteristics columns appear to differ in many cases.

Note fully duplicate records have previously been dropped in `concat/src/concat.r`.

```{r duplicate_placements}

# Flag all duplicate placement book-ins within each stay
df <- df %>% 
  group_by(stayid) %>% 
  mutate(is_dupe_book_in = duplicated(detention_book_in_date_and_time) |
           duplicated(detention_book_in_date_and_time, fromLast = TRUE),
         is_dupe_book_out = duplicated(detention_book_out_date_time) |
            duplicated(detention_book_out_date_time, fromLast = TRUE)) %>% 
  ungroup()

h1 <- df %>% 
  filter(is_dupe_book_in == TRUE) %>% 
  ggplot(aes(x = placement_length_min)) +
  geom_histogram()

h1

```

There are many zero length placements that are not associated with duplicate placement records.

Very small number of duplicate placement records have non-zero placement length (these records may generate double count in `headcount/src/headcount.R`.)

```{r}
 
# Check if all zero length placements are duplicates (probably not)
dupeplacements <- df[df$is_dupe_book_in == TRUE, "recid"]
dupeplacementstays <- unique(df[df$is_dupe_book_in == TRUE, "stayid"])

zeroplacements <- df[df$placement_length_min == 0, "recid"]
zeroplacementstays <- unique(df[df$placement_length_min == 0, "stayid"])

# Stays with non zero-length duplicate placements
non_zero_dupe_stayids <- setdiff(unlist(dupeplacementstays), unlist(zeroplacementstays))

non_zero_dupe_stays <- df %>%
  filter(stayid %in% non_zero_dupe_stayids)

```

# Date range

The following table offers an overview of placement and stay book-in and book-out fields. Note this dataset should include all detention records for all people who experienced ICE detention during October 1, 2011 through January 4, 2024; accordingly, this includes people with placements and stays beginning prior to October 1, 2011.

```{r date_range}

date_cols <- c("stay_book_in_date_time",
               "detention_book_in_date_and_time",
               "detention_book_out_date_time",
               "stay_book_out_date_time",
               "detention_release_reason",
               "stay_release_reason")

skimr::skim(df[date_cols])

unique_stays <- df %>% 
  distinct(stayid, .keep_all=TRUE)

skimr::skim(unique_stays[date_cols])

```

# Incomplete placements/stays

We have interpreted records with null placement and stay book-out values as representing ongoing detentions at the time of generation of the dataset. 

It is logical to observe more null stay book-out values than null placement book-out dates, as each "bloc" of successive placement records associated with an ongoing stay will be present multiple null stay book-out values.

We should not expect to see any records with null placement book-out and non-null stay book-out: `r stopifnot(nrow(df[is.na(df$detention_book_out_date_time) & !is.na(df$stay_book_out_date_time),]) == 0)`.

However, we do see records with no null placement book-out values, but null stay book-out values, indicating that the stay appears to be ongoing but no ongoing placement appears in the dataset. These records will not confound detained population calculations in `headcount/src/headcount.R`.

```{r}

unique_stays <- df %>% 
  distinct(stayid, .keep_all = TRUE)

stopifnot(unique(unique_stays$placement_count) == 1)

test <- unique_stays %>% 
  filter(is.na(stay_book_out_date_time),
         !is.na(detention_book_out_date_time))

test_ids <- test$stayid

test2 <- df %>% 
  filter(stayid %in% test_ids)

```

# Placement length

`r sum(df$placement_length == 0, na.rm = TRUE)` completed placements have a `placement_length` of zero; `r sum(df$placement_length < 0, na.rm = TRUE)` have negative `placement_length` values.

Placements with length zero will not contribute to headcount calculations in `headcount/src/headcount.R` unless book in time is exactly 01:00 hours UTC; this amounts to very few records at any given time.

Negative length placements will not contribute to headcount calculations.

```{r placement_length}

dat <- df %>% 
  mutate(placement_length_bin = cut(placement_length,
                           breaks = c(-Inf, 0, 3, 30, 90, 360, Inf),
                           right = TRUE))
p1 <- dat %>%
  count(placement_length_bin) %>% 
  ggplot(aes(x = placement_length_bin, y = n)) +
  geom_col()

p1

```

# Stay length

`r sum(unique_stays$stay_length == 0, na.rm = TRUE)` completed stays have a `stay_length` of zero; `r sum(unique_stays$stay_length < 0, na.rm = TRUE)` have negative `stay_length` values. 

```{r stay_length}

dat <- unique_stays %>%
  mutate(stay_length_bin = cut(placement_length,
                           breaks = c(-Inf, 0, 3, 30, 90, 360, Inf),
                           right = TRUE))

p1 <- dat %>%
  count(stay_length_bin) %>% 
  ggplot(aes(x = stay_length_bin, y = n)) +
  geom_col()

p1

```

# Successive placements

Examining characteristics of successive placements, using FY2020 as test case, as this calculation is memory-intensive for full dataset. Some detention placement book-ins have exact same time as previous placement book-out. Others may have book-in prior to previous book-out, or with long gaps between placements.

```{r placement_lag}

test_df <- df %>% 
  filter(detention_book_in_date_and_time >= '2019-10-01',
         detention_book_in_date_and_time <= '2020-09-30')

test_df <- test_df %>%
  group_by(stayid) %>%
  mutate(bookout_lag = lag(detention_book_out_date_time, n=1, order_by=stayid),
         bookin_diff = difftime(detention_book_in_date_and_time, bookout_lag, unit="days")) %>%
  ungroup()

dat <- as.numeric(unlist(test_df[!is.na(test_df$bookin_diff), 'bookin_diff']))

hist(dat, right = T, breaks=c(-100, -1, 0, 1, 250))

```

```{r}

sample_cols <- c("stay_book_in_date_time",
                 "detention_book_in_date_and_time", # Note irregular field name format
                 "detention_book_out_date_time",
                 "stay_book_out_date_time",
                 "citizenship_country",
                 "gender",
                 "birth_year",
                 "detention_facility",
                 "area_of_responsibility",
                 "detention_release_reason",
                 "stay_release_reason",
                 "anonymized_identifier",
                 "stay_length",
                 "placement_length",
                 "bookout_lag",
                 "bookin_diff")

ex_zero <- test_df[test_df$bookin_diff == 0,] %>%
  filter(!is.na(stay_book_in_date_time))

ex_zero_ids <- unique(ex_zero$anonymized_identifier)

df_ex_zero <- test_df %>%
  filter(anonymized_identifier %in% ex_zero_ids) %>% 
  dplyr::select(all_of(sample_cols))


###

ex_neg <- test_df[test_df$bookin_diff < 0,] %>%
  filter(!is.na(stay_book_in_date_time))

ex_neg_ids <- unique(ex_neg$anonymized_identifier)

df_ex_neg <- test_df %>%
  filter(anonymized_identifier %in% ex_neg_ids) %>% 
  dplyr::select(all_of(sample_cols))

###

ex_gap <- test_df[test_df$bookin_diff > 3,] %>%
  filter(!is.na(stay_book_in_date_time))

ex_gap_ids <- unique(ex_gap$anonymized_identifier)

df_ex_gap <- test_df %>%
  filter(anonymized_identifier %in% ex_gap_ids) %>% 
  dplyr::select(all_of(sample_cols))

```


Examining stay blocs with odd placement timings (instant transfers, impossible placement lengths, long gaps between placements).

Many (but not all) records involving instant transfers (book-out time exactly equal to subsequent book-in time) appear to involve cases with `detention_release_reason` values of "Processing Disposition Changed Locally"; this would make sense if these placements are "on paper only" or simply administrative changes. These cases are not uncommon (`r length(ex_zero_ids)` individual stays during FY2020), so this is a good argument for using a headcount methodology which is not inclusive of book-out dates in order to not double-count these placements (see below).

Negative placement gaps mean that the dataset records an individual as being in two places at one time, e.g. one placement is not completed but subsequent has already started. These are not common (`r length(ex_neg_ids)` individual stays during FY2020). These records will result in double counting in detention headcounts.

# Exclusive versus inclusive headcount methods

Comparing outcomes of headcount based on inclusive versus exclusive indexing of placement book-out date-time. Exclusive indexing is more conservative; inclusive indexing may over-count if successive placements have exact same book-out and book-in values.

```{r daily_headcount, message=FALSE, warning=FALSE, include=TRUE}

# Date exclusive indexing (default output of `headcount/src/headcount.R`)
headcount_exc <- read_delim(here::here('export', 'output', 'headcount_fy12-24ytd.csv.gz'), delim='|')

# Date inclusive indexing
headcount_inc <- read_delim(here::here('headcount', 'frozen', 'inc_headcount_fy12-24ytd.csv.gz'), delim='|')

# Exclusive of book-out date-time
headcount_exc <- headcount_exc %>% 
  mutate(date = as.Date(date),
         v = "exclusive")

# Inclusive of book-out date-time
headcount_inc <- headcount_inc %>% 
  mutate(date = as.Date(date),
         v = "inclusive")

hc_compare <- rbind(headcount_exc, headcount_inc)

p1 <- hc_compare %>% 
  group_by(date, v) %>% 
  summarize(n = sum(n)) %>% 
  ggplot(aes(x = date, y = n, color = v)) +
  geom_line()

p1

dat <- hc_compare %>% 
  group_by(date, v) %>% 
  summarize(n = sum(n)) %>% 
  pivot_wider(id_cols = date, names_from = v, values_from = n) %>% 
  mutate(diff = exclusive - inclusive)

p2 <- dat %>% 
  ggplot(aes(x = date, y = diff)) +
  geom_line()

p2

```

# Difference between stay/placement book in/out times

Detention stay book in/book out values do not always coincide with associated initial/final detention placement records. This is most common in data for placements in FY2011 and prior, but occurs throughout dataset.

# Difference between stay/placement headcounts

As a result of this and gaps between placements, a detention headcount based on stay book in/out values results in higher approximate total detained population, sometimes by a significant margin.

```{r stay_headcounter, message=FALSE, warning=FALSE, include=TRUE}

# Transform data to "fill in" missing `detention_book_out_date_time` values
# in order to account for ongong detention stays at time of release of data
max_date <- max(df$stay_book_out_date_time, na.rm=TRUE)

unique_stays <- unique_stays %>%
  mutate(stay_book_out_date_time_min = 
           case_when(is.na(stay_book_out_date_time) ~ max_date,
                     TRUE ~ stay_book_out_date_time))

# Define timeline for calculation of daily detained population
timeline_start <- min(unique_stays$stay_book_in_date_time, na.rm=TRUE)
timeline_end <- max(unique_stays$stay_book_out_date_time, na.rm=TRUE)
timeline <- seq(timeline_start, timeline_end, by='day')

# Function counts all in-range detention placement records in dataset `df` for a
# given `date` by a given grouping variable `var`
stay_headcounter <- function(date, df, group_vars) {
  
  in_range <- df[df$stay_book_in_date_time <= date & df$stay_book_out_date_time_min >= date,]
  
  in_range %>% 
    group_by(across(all_of(group_vars))) %>% 
    summarize(n = n()) %>% 
    complete(fill = list(n = 0)) %>% 
    mutate(date=date)
  
  }


# Apply function to timeline
stay_headcount <- lapply(timeline, stay_headcounter, df=unique_stays, group_vars=c('detention_facility_code'))

# Transform output into data frame
stay_headcount_data <- map_dfr(stay_headcount, bind_rows)

write_delim(stay_headcount_data, here::here('analyze', 'frozen', 'stay_headcount_fy12-24ytd.csv.gz'), delim='|')

```

```{r compare_stay_placement_headcounts}

stay_headcount_data <- read_delim(here::here('analyze', 'frozen', 'stay_headcount_fy12-24ytd.csv.gz'), delim='|')


# Plot headcount
p1 <- stay_headcount_data %>%
  group_by(date) %>% 
  summarize(n = sum(n)) %>% 
  filter(date >= "2010-10-01") %>%
  ggplot(aes(x = date, y = n) ) +
  geom_line() +
  labs(title = "Daily detained population")

p1

stay_headcount_daily <- stay_headcount_data %>%
  group_by(date) %>% 
  summarize(n = sum(n)) %>% 
  mutate(v = "stays",
         date = as.Date(date))

headcount_daily <- headcount_exc %>%
  group_by(date) %>% 
  summarize(n = sum(n)) %>% 
  mutate(v = "placements",
         date = as.Date(date))

dat <- rbind(stay_headcount_daily, headcount_daily)

p2 <- dat %>% 
  ggplot(aes(x = date, y = n, color = v)) +
  geom_line()

p2

dat_wide <- dat %>% 
  pivot_wider(id_cols = date, names_from = v, values_from = n) %>% 
  mutate(diff = stays - placements)

p3 <- dat_wide %>% 
  ggplot(aes(x = date, y = diff)) +
  geom_line()

p3

```

# Headcount of zero length placements

As currently implemented, zero-length placements will only contribute to headcount if placement book in time is at exactly 01:00 hours UTC.

```{r headcounter, message=FALSE, warning=FALSE, include=TRUE}

# Transform data to "fill in" missing `detention_book_out_date_time` values
# in order to account for ongong detention stays at time of release of data
max_date <- max(df$stay_book_out_date_time, na.rm=TRUE)

df <- df %>%
  mutate(detention_book_out_date_time_min = 
           case_when(is.na(detention_book_out_date_time) ~ max_date,
                     TRUE ~ detention_book_out_date_time))

# Define timeline for calculation of daily detained population
timeline_start <- min(df$stay_book_in_date_time, na.rm=TRUE)
timeline_end <- max(df$stay_book_out_date_time, na.rm=TRUE)
timeline <- seq(timeline_start, timeline_end, by='day')

# Function counts all in-range detention placement records in dataset `df` for a
# given `date` by a given grouping variable `var`
headcounter <- function(date, df, group_vars) {
  
  in_range <- df[df$detention_book_in_date_and_time <= date & df$detention_book_out_date_time_min >= date,]
  
  in_range %>% 
    group_by(across(all_of(group_vars))) %>% 
    summarize(n = n()) %>% 
    complete(fill = list(n = 0)) %>% 
    mutate(date=date)
  
  }

# Generate limited sample dataset
test_df <- df %>%
  filter(placement_length_min == 0)

# Apply function to timeline
example_headcount <- lapply(timeline, headcounter, df=test_df, group_vars=c('recid'))

# Transform output into data frame
example_headcount_data <- map_dfr(example_headcount, bind_rows)

# # Plot headcount
p1 <- example_headcount_data %>%
  filter(date >= "2010-10-01") %>%
  ggplot(aes(x = date, y = n) ) +
  geom_area() +
  labs(title = "Daily detained population")

p1

```
