---
title: "FY11-24YTD ICE detention headcount facility-level analysis"
author: "UWCHR"
date: "2025-01-31"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

```{r import_data}

library(pacman)
p_load(tidyverse, arrow, lubridate, zoo, digest, ggplot2, plotly, gghighlight)

headcount <- read_delim(here::here('export', 'output', 'headcount_fy12-24ytd.csv.gz'), delim='|')

df <- data.table::fread(file = here::here('export', 'output', 'ice_detentions_fy12-24ytd.csv.gz'),
                        sep = '|')

data.table::setDF(df)

detloc_aor <- df %>% 
  distinct(detention_facility_code, area_of_responsibility) %>% 
  arrange(detention_facility_code, area_of_responsibility)

df <- df %>% 
  mutate(arresting_authority = case_when(
    str_detect(final_program, "287G Program") ~ "ICE",
    str_detect(final_program, "ERO Criminal Alien Program") ~ "ICE",
    str_detect(final_program, "Border Patrol") ~ "USBP",
    str_detect(final_program, "Fugitive Operations") ~ "ICE",
    str_detect(final_program, "Violent Criminal Alien Section") ~ "ICE",
    str_detect(final_program, "Law Enforcement Area Response Units") ~ "ICE",
    str_detect(final_program, "Non-User Fee Investigations") ~ "Other",
    str_detect(final_program, "Inspections - Land") ~ "CBP",
    str_detect(final_program, "Detention and Deportation") ~ "ICE",
    str_detect(final_program, "Intelligence") ~ "ICE",
    str_detect(final_program, "Detained Docket Control") ~ "ICE",
    str_detect(final_program, "Non-Detained Docket Control") ~ "ICE",
    str_detect(final_program, "Inspections - Air") ~ "CBP",
    str_detect(final_program, "Asylum") ~ "ICE",
    str_detect(final_program, "Joint Criminal Alien Response Team") ~ "ICE",
    str_detect(final_program, "Alternatives to Detention") ~ "ICE",
    str_detect(final_program, "Inspections - Sea") ~ "CBP",
    str_detect(final_program, "PICS Default Value - for user initialization only") ~ "Other",
    str_detect(final_program, "Adjudications") ~ "ICE",
    str_detect(final_program, "User Fee Investigations") ~ "Other",
    str_detect(final_program, "Joint Terrorism Task Force") ~ "ICE",
    str_detect(final_program, "Examinations") ~ "ICE",
    str_detect(final_program, "Quick Response Team") ~ "ICE",
    str_detect(final_program, "Default program area for interface records") ~ "Other",
    str_detect(final_program, "Probation and Parole") ~ "ICE",
    str_detect(final_program, "HSI Criminal Arrest Only") ~ "ICE",
    str_detect(final_program, "Juvenile") ~ "ICE",
    str_detect(final_program, "Mobile Criminal Alien Team") ~ "ICE",
    str_detect(final_program, "ERO Criminal Prosecutions") ~ "ICE"
  ))

headcount <- left_join(headcount, detloc_aor, by="detention_facility_code")

```

```{r facil_setup}
# Get records for specific facility:
facil_code <- "CSCNWWA"
facil_name <- unique(df[df$detention_facility_code == facil_code, "detention_facility"])
facil_aor <- unique(df[df$detention_facility_code == facil_code, "area_of_responsibility"])

# Get all anon ID values associated with facility
facil_ids <- df %>% 
  filter(detention_facility_code == facil_code) %>% 
  distinct(anonymized_identifier)

# Get all records for these anon IDs at facility
facil_records <- df %>% 
  filter(anonymized_identifier %in% unlist(facil_ids),
         detention_facility_code == facil_code)

# Get distinct record for each stay by person at facility
facil_unique_stays <- facil_records %>% 
  distinct(anonymized_identifier, stay_count, .keep_all = TRUE)

```

## Specific facility: `r facil_name`

```{r facil_headcount}

p1 <- headcount %>% 
  filter(date >= "2011-10-01",
         detention_facility_code == facil_code) %>% 
  ggplot(aes(x = as.Date(date), y = n)) +
  geom_line() +
  ylim(0, NA) +
  ylab("Pop.") +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  labs(title = "Daily detained population, FY2012-Jan. 2024",
       subtitle = paste0(facil_name))

p1

```

```{r det_rank}

adp_fy <- headcount %>% 
  filter(date >= "2011-10-01") %>% 
  mutate(fy = as.factor(substr(quarter(date, fiscal_start=10, type="year.quarter"), 1,4))) %>% 
  group_by(fy, detention_facility_code) %>% 
  summarize(adp = mean(n, na.rm=TRUE))

active_facil_fy <- headcount %>% 
  filter(date >= "2011-10-01") %>% 
  mutate(fy = as.factor(substr(quarter(date, fiscal_start=10, type="year.quarter"), 1,4))) %>% 
  group_by(fy, detention_facility_code) %>% 
  summarize(adp = mean(n, na.rm=TRUE)) %>% 
  filter(adp > 0) %>% 
  group_by(fy) %>% 
  summarize(active_facil = n_distinct(detention_facility_code))

adp_rank <- adp_fy %>% 
  left_join(active_facil_fy, by = 'fy') %>% 
  arrange(fy, desc(adp)) %>% 
  group_by(fy) %>% 
  mutate(ranking = row_number()) %>% 
  ungroup()

```

With an average detained population of `r as.numeric(adp_rank[adp_rank$detention_facility_code == facil_code & adp_rank$fy == 2023, "adp"])` during FY 2023, `r facil_name` ranked `r as.numeric(adp_rank[adp_rank$detention_facility_code == facil_code & adp_rank$fy == 2023, "ranking"])` out of `r as.numeric(adp_rank[adp_rank$detention_facility_code == facil_code & adp_rank$fy == 2023, "active_facil"])` active detention facilities.

```{r det_compare_natl}

# National daily pop counts
p1 <- headcount %>% 
  filter(date >= "2011-10-01") %>% 
  group_by(date, detention_facility_code) %>% 
  summarize(total_detained = sum(n)) %>% 
  ggplot(aes(x = as.Date(date), y = total_detained, color=detention_facility_code, group=detention_facility_code)) +
  geom_line() +
  gghighlight(detention_facility_code == facil_code) +
  ylim(0, NA) +
  ylab("Pop.") +
  xlab("Date") +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  labs(title = "National daily detained population, FY2012-Jan. 2024",
       subtitle = paste0(facil_name, " highlighted"))

p1

```

```{r det_compare_aor}

p1 <- headcount %>% 
  filter(date >= "2011-10-01",
         area_of_responsibility == facil_aor) %>% 
  group_by(date, detention_facility_code) %>% 
  summarize(total_detained = sum(n)) %>% 
  ggplot(aes(x = as.Date(date), y = total_detained, color=detention_facility_code, group=detention_facility_code)) +
  geom_line() +
  gghighlight(detention_facility_code == facil_code) +
  ylim(0, NA) +
  ylab("Pop.") +
  xlab("Date") +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  # scale_y_log10() +
  labs(title = paste0(facil_aor, " daily detained population, FY2012-Jan. 2024"),
       subtitle = paste0(facil_name, " highlighted"))

p1

p2 <- headcount %>% 
  filter(date >= "2011-10-01",
         area_of_responsibility == facil_aor) %>% 
  mutate(detention_facility_code = case_when(detention_facility_code == facil_code ~ facil_code,
                                             TRUE ~ "ALL OTHERS")) %>% 
  group_by(date, detention_facility_code) %>% 
  summarize(total_detained = sum(n)) %>% 
  ggplot(aes(x = as.Date(date), y = total_detained, color=detention_facility_code, group=detention_facility_code)) +
  geom_line() +
  ylim(0, NA) +
  ylab("Pop.") +
  xlab("Date") +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  # scale_y_log10() +
  labs(title = paste0(facil_aor, " daily detained population, FY2012-Jan. 2024"),
       subtitle = paste0(facil_name, " versus all others"))

p2

```

```{r det_aor_total}

p1 <- headcount %>% 
  filter(date >= "2011-10-01",
         area_of_responsibility == facil_aor) %>% 
  group_by(date) %>% 
  summarize(total_detained = sum(n)) %>% 
  mutate(year_mth = as.yearmon(date)) %>% 
  group_by(year_mth) %>% 
  summarize(adp = mean(total_detained, na.rm=TRUE)) %>% 
  ggplot(aes(x = year_mth, y = adp)) +
  geom_line() +
  ylim(0, NA) +
  ylab("ADP") +
  xlab("") +
  labs(title = paste0(facil_aor, " avg. daily detained population, FY2012-Jan. 2024"))

p1

```

```{r atd}

# trac_atd <- read_delim(here::here('analyze', 'input', 'trac_atd_2024-02-24.csv'), delim=',')
# 
# names(trac_atd) <- tolower(names(trac_atd))
# names(trac_atd) <- str_replace_all(names(trac_atd), " ", "_")
# 
# trac_atd$Date <- as.Date(trac_atd$date, format = "%m/%d/%Y")
# trac_atd$year_mth <- as.yearmon(trac_atd$Date)
# 
# aor_atd <- trac_atd %>% 
#   filter(!is.na(ice_area_of_responsibility),
#          atd_technology == "All",
#          ice_area_of_responsibility == "Seattle")
# 
# # Remove outlier
# aor_atd <- aor_atd[!aor_atd$number_under_atd %in% boxplot.stats(aor_atd$number_under_atd)$out,]
# 
# monthly_aor_atd <- aor_atd %>% 
#   group_by(year_mth) %>% 
#   summarize(avg_atd = mean(number_under_atd, na.rm = FALSE)) %>% 
#   mutate(type = "atd") %>% 
#   rename(n = avg_atd)
# 
# monthly_adp <- headcount %>% 
#   filter(date >= "2011-10-01",
#          area_of_responsibility == facil_aor) %>% 
#   group_by(date) %>% 
#   summarize(total_detained = sum(n)) %>% 
#   mutate(year_mth = as.yearmon(date)) %>% 
#   group_by(year_mth) %>% 
#   summarize(adp = mean(total_detained, na.rm=TRUE)) %>% 
#   mutate(type = "adp") %>% 
#   rename(n = adp)
# 
# dat <- rbind(monthly_aor_atd, monthly_adp)
# 
# dat %>% ggplot(aes(x = year_mth, y = n, color = type, group = type)) +
#   geom_line() +
#   labs(title = "Avg. detained pop. + Alternatives to detention",
#        subtitle = facil_aor)

```

```{r age}

# Includes some invalid birth years, e.g. 1900

facil_unique_stays <- facil_unique_stays %>% 
  mutate(detention_cy = substr(quarter(detention_book_in_date_and_time, type="year.quarter"), 1, 4),
         age = as.numeric(detention_cy) - birth_year,
         age_group = cut(age,
                         breaks = c(0, 18, 25, 40, 60, Inf),
                         right = FALSE))

hist(facil_unique_stays$birth_year)
hist(facil_unique_stays$age)

p1 <- facil_unique_stays %>% 
  mutate(detention_fy = substr(quarter(detention_book_in_date_and_time, fiscal_start = 10, type="year.quarter"), 1, 4)) %>%
  count(age_group, detention_fy) %>% 
  filter(detention_fy >= 2012) %>% 
  ggplot(aes(x = detention_fy, y = n, fill = age_group)) +
  geom_col(position="fill")
  
p1

p2 <- facil_unique_stays %>% 
  mutate(detention_fy = substr(quarter(detention_book_in_date_and_time, fiscal_start = 10, type="year.quarter"), 1, 4)) %>%
  group_by(detention_fy) %>% 
  summarize(avg_age = mean(age, na.rm=TRUE)) %>% 
  filter(detention_fy >= 2012) %>% 
  ggplot(aes(x = detention_fy, y = avg_age)) +
  geom_col() +
  labs(title = "Average approx. age of detained pop.")
  
p2

```

```{r facil_age_vs_natl}

unique_stays <- df %>%
  distinct(stayid, .keep_all = TRUE) %>% 
  mutate(detention_cy = substr(quarter(detention_book_in_date_and_time, type="year.quarter"), 1, 4),
         age = as.numeric(detention_cy) - birth_year,
         age_group = cut(age,
                         breaks = c(0, 18, 25, 40, 60, Inf),
                         right = FALSE))

dat1 <- facil_unique_stays %>% 
  mutate(detention_fy = substr(quarter(detention_book_in_date_and_time, fiscal_start = 10, type="year.quarter"), 1, 4)) %>%
  count(age_group, detention_fy) %>% 
  filter(detention_fy >= 2012) %>% 
  mutate(group = facil_code)

dat2 <- unique_stays %>% 
  mutate(detention_fy = substr(quarter(detention_book_in_date_and_time, fiscal_start = 10, type="year.quarter"), 1, 4)) %>%
  count(age_group, detention_fy) %>% 
  filter(detention_fy >= 2012) %>% 
  mutate(group = "Natl")

dat <- rbind(dat1, dat2)

p1 <- dat %>%
  ggplot(aes(x = detention_fy, y = n, fill = age_group)) +
  geom_col(position="fill") + 
  facet_wrap(~group)
  
p1

p2 <- dat %>%
  filter(age_group != "[0,18)") %>% 
  ggplot(aes(x = detention_fy, y = n, fill = age_group)) +
  geom_col(position="fill") + 
  facet_wrap(~group)
  
p2

```

```{r gender}

dat1 <- facil_unique_stays %>% 
  mutate(detention_fy = substr(quarter(detention_book_in_date_and_time, fiscal_start = 10, type="year.quarter"), 1, 4)) %>%
  count(gender, detention_fy) %>% 
  filter(detention_fy >= 2012) %>% 
  mutate(group = facil_code)

dat2 <- unique_stays %>% 
  mutate(detention_fy = substr(quarter(detention_book_in_date_and_time, fiscal_start = 10, type="year.quarter"), 1, 4)) %>%
  count(gender, detention_fy) %>% 
  filter(detention_fy >= 2012) %>% 
  mutate(group = "Natl")


dat <- rbind(dat1, dat2)

p1 <- dat %>% 
  ggplot(aes(x = detention_fy, y = n, fill = gender, group = group)) +
  geom_col(position='fill') + 
  facet_wrap(~group)

p1

```


```{r bookin_facil}

# Book-in facility
facil_unique_stays %>% 
  count(first_facil) %>% 
  arrange(desc(n))

top_bookin_facil <- facil_unique_stays %>% 
  filter(stay_book_in_date_time >= "2011-10-01") %>% 
  count(first_facil) %>% 
  arrange(desc(n)) %>% 
  head(15)

p1 <- facil_unique_stays %>% 
  filter(stay_book_in_date_time >= "2011-10-01") %>% 
  mutate(detention_fy = str_sub(quarter(stay_book_in_date_time,
                                with_year = TRUE,
                                fiscal_start = 10), 1, 4),
         first_facil = case_when(first_facil %in% unlist(top_bookin_facil$first_facil) ~ first_facil,
                                 TRUE ~ "ALL OTHERS")) %>% 
  count(detention_fy, first_facil) %>%
  ggplot(aes(x = detention_fy, y = n, fill=first_facil)) +
  geom_col(position="fill")

ggplotly(p1)

```

```{r bookin_facil_aor}

facil_unique_stays <- left_join(facil_unique_stays, detloc_aor, by=c("first_facil" = "detention_facility_code"), suffix = c("", "_first_facil"))

dat <- facil_unique_stays %>% 
  distinct(anonymized_identifier, .keep_all = TRUE) %>% 
  filter(stay_book_in_date_time >= "2011-10-01") %>% 
  mutate(detention_fy = str_sub(quarter(stay_book_in_date_time,
                                with_year = TRUE,
                                fiscal_start = 10), 1, 4)) %>%
  count(detention_fy, first_aor)

top_bookin_aors <- facil_unique_stays %>% 
  count(first_aor) %>% 
  arrange(desc(n))

p1 <- dat %>% 
  mutate(first_aor = case_when(first_aor %in% head(top_bookin_aors$first_aor, 5) ~ first_aor,
         TRUE ~ "All others"),
    first_aor = str_replace_all(first_aor, " Area of Responsibility", "")) %>% 
  ggplot(aes(x = detention_fy, y = n, color=first_aor, fill=first_aor, group = first_aor)) +
  geom_col() +
  labs(title = "Initial book-in AOR",
       subtitle = paste0("Individuals with stays at ", facil_name))

p1

p2 <- dat %>% 
  mutate(first_aor = str_replace_all(first_aor, " Area of Responsibility", "")) %>%
  ggplot(aes(x = detention_fy, y = n, color=first_aor, fill=first_aor, group = first_aor)) +
  geom_col() +
  scale_x_discrete(breaks=seq(2012, 2024, 4)) +
  facet_wrap(~first_aor) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Initial book-in AOR",
       subtitle = paste0("Individuals with stays at ", facil_name))
 
p2

```

```{r los}

# LOS groups
facil_los_groups <- facil_unique_stays %>%
  dplyr::select(stay_length) %>% 
  mutate(los_g = cut(as.numeric(stay_length, unit='days'),
                           breaks = c(-Inf, 3, 15, 30, 90, 180, 365, Inf),
                           right = FALSE)) %>% 
  count(los_g) %>% 
  mutate(prop = n / sum(n)) %>% 
  ungroup()

natl_los_groups <- unique_stays %>% 
  dplyr::select(stay_length) %>% 
  mutate(los_g = cut(as.numeric(stay_length, unit='days'),
                           breaks = c(-Inf, 3, 15, 30, 90, 180, 365, Inf),
                           right = FALSE)) %>% 
  count(los_g) %>% 
  mutate(prop = n / sum(n)) %>% 
  ungroup()


dat <- left_join(natl_los_groups, facil_los_groups, by='los_g', suffix=c(".natl", ".facil")) %>% 
  pivot_longer(cols=-c('los_g')) %>%
  filter(str_detect(name, "prop")) %>% 
  mutate(value = case_when(name == "prop.natl" ~ -value,
                   name == "prop.facil" ~ value))

los_prop_range <- range(dat$value)
los_prop_range_breaks <- pretty(dat$value, 10)

p1 <- dat %>% 
  filter(!is.na(los_g)) %>% 
  ggplot(aes(x = value, y = los_g, fill=name, color=name)) +
    geom_col() +
    scale_y_discrete(limits=rev) +
    scale_x_continuous(
      breaks = los_prop_range_breaks,
      labels = function(x) scales::percent(abs(x))) +
  labs(title = "Length of stay distribution",
       subtitle = paste0(facil_name, " vs. national"),
       caption = "* Note this comparison includes both under-72 and over-72 hour facilities.")

p1


```

```{r alos_trend}

# ALOS over time
p1 <- facil_unique_stays %>% 
  distinct(anonymized_identifier, .keep_all=TRUE) %>% 
  filter(stay_book_in_date_time >= "2011-10-01",
         stay_book_in_date_time <= "2023-09-30") %>% 
  mutate(detention_fy = substr(quarter(stay_book_in_date_time, with_year = TRUE, fiscal_start = 10), 1,4)) %>% 
  group_by(detention_fy) %>% 
  summarize(alos = mean(as.numeric(stay_length, unit='days'), na.rm=TRUE),
            alos_min = mean(as.numeric(stay_length_min, unit='days'))) %>% 
  pivot_longer(cols=-detention_fy) %>% 
  ggplot(aes(x = detention_fy, y = value, color = name, group=name)) +
  geom_line() +
  ylim(0, NA) +
  labs(title = "Average length of stay",
       subtitle = paste0("Individuals detained at ", facil_name))

p1

```

```{r alos_compare}

# Facil ALOS versus national ALOS

natl_alos <- unique_stays %>% 
  filter(stay_book_in_date_time >= "2011-10-01") %>% 
  mutate(detention_qtr = quarter(stay_book_in_date_time, with_year = TRUE, fiscal_start = 10)) %>% 
  group_by(detention_qtr) %>% 
  summarize(alos = mean(as.numeric(stay_length, units='days'), na.rm=TRUE))

# The following includes a quick and dirty method for excluding under 72-hour facilities; but should do a better version of this

facil_alos <- unique_stays %>% 
  filter(stay_book_in_date_time >= "2011-10-01") %>% 
  mutate(detention_qtr = quarter(stay_book_in_date_time, with_year = TRUE, fiscal_start = 10)) %>% 
  group_by(detention_qtr, detention_facility_code) %>% 
  summarize(alos = mean(as.numeric(stay_length, units='days'), na.rm=TRUE),
            med_alos = median(as.numeric(stay_length, units='days'), na.rm=TRUE)) %>% 
  mutate(under_72 = med_alos < 3) %>% 
  filter(under_72 == FALSE)

dat <- facil_alos %>% 
  dplyr::select(-c(med_alos, under_72)) %>% 
  left_join(natl_alos, by = 'detention_qtr', suffix=c(".facil", ".natl")) %>% 
  rename(national_alos = alos.natl,
         facility_alos = alos.facil) %>% 
  pivot_longer(cols=-c('detention_qtr', 'detention_facility_code')) %>% 
  filter(detention_facility_code == facil_code)

p1 <- dat %>% 
  ggplot(aes(x=as.factor(detention_qtr), y = value, color = name, group = name)) +
  geom_line() +
  scale_x_discrete(breaks = levels(as.factor(dat$detention_qtr))[c(T, rep(F, 3))],
                   labels = function(x) substr(as.character(x), 1, 4)) +
  ylim(0, NA) +
  ylab("Days") +
  xlab("Detention quarter") +
  labs(title = "Comparison of quarterly Average Length of Stay (ALOS)",
       subtitle = paste0("Nat'l vs. ", facil_name))

p1

```

```{r stays}

# Multiple stay trends
# Multiple placement/transfer trends !DOUBLE CHECK THIS LOGIC!
# Note we have a few variables to work with here:
# `total_stays`: number of unique detention stay book in values per anon ID
# `total_placements`: number of unique detention placement book in values per anon ID
# `stay_placements`: number of unique detention placement book in values per stay per anon ID

p3 <- facil_unique_stays %>% 
  filter(stay_book_in_date_time > "2011-10-01",
         !duplicated(anonymized_identifier)) %>% 
  mutate(detention_fy = str_sub(quarter(stay_book_in_date_time,
                                with_year = TRUE,
                                fiscal_start = 10), 1, 4)) %>% 
  group_by(detention_fy) %>% 
  summarize(avg_stays = mean(total_stays, na.rm=TRUE),
            avg_placements = mean(stay_placements, na.rm=TRUE)) %>%
  pivot_longer(cols=-detention_fy) %>% 
  ggplot(aes(x=detention_fy, y=value, fill=name, group=name)) +
  geom_col(position='dodge')

p3

```

```{r alos_compare}

national_alos <- unique_stays %>% 
  filter(stay_book_in_date_time > "2011-10-01",
         stay_book_in_date_time <= "2023-09-30") %>% 
  mutate(detention_fy = substr(quarter(stay_book_in_date_time, with_year = TRUE, fiscal_start = 10), 1, 4)) %>% 
  group_by(detention_fy) %>% 
  summarize(alos = mean(as.numeric(stay_length, unit='days'), na.rm=TRUE),
            alos_min = mean(as.numeric(stay_length_min, unit='days')))

facil_alos <- facil_unique_stays %>% 
  filter(stay_book_in_date_time >= "2011-10-01",
         stay_book_in_date_time <= "2023-09-30") %>% 
  mutate(detention_fy = substr(quarter(stay_book_in_date_time, with_year = TRUE, fiscal_start = 10), 1, 4)) %>% 
  group_by(detention_fy) %>% 
  summarize(alos = mean(as.numeric(stay_length, unit='days'), na.rm=TRUE),
            alos_min = mean(as.numeric(stay_length_min, unit='days'))) %>% 
  rename(facil_alos = alos,
         facil_alos_min = alos_min)

dat <- left_join(national_alos, facil_alos, by='detention_fy')

p1 <- dat %>% 
  dplyr::select(detention_fy, alos, facil_alos) %>% 
  pivot_longer(cols=-c('detention_fy')) %>% 
  ggplot(aes(x=as.factor(detention_fy), y=value, color=name, group=name)) + 
  geom_line() +
  ylim(0, NA)

p1  

p2 <- dat %>%
  mutate(facil_alos_ratio = facil_alos / alos) %>%
  pivot_longer(cols=-c('detention_fy')) %>% 
  filter(name == "facil_alos_ratio") %>% 
  ggplot(aes(x=as.factor(detention_fy), y=value, color=name, fill=name, group=name)) + 
  geom_col()

p2

```

## Current detained pop

```{r facil_current_detained}

all_current <- unique_stays %>% 
  filter(current_stay == TRUE) %>% 
  distinct(anonymized_identifier, .keep_all = TRUE)

stopifnot(length(unique(all_current$anonymized_identifier)) == nrow(all_current))

facil_current <- facil_unique_stays %>% 
  filter(current_stay == TRUE)

stopifnot(length(unique(facil_current$anonymized_identifier)) == nrow(facil_current))

top_cit <- all_current %>% 
  filter(detention_facility_code == facil_code) %>% 
  count(citizenship_country) %>% 
  arrange(desc(n)) %>% 
  head(10)

all_current %>% 
  filter(detention_facility_code == facil_code) %>%  
  count(gender) %>% 
  arrange(desc(n))

all_current %>% 
  filter(detention_facility_code == facil_code) %>%  
  count(`most_serious_conviction_(msc)_criminal_charge_category`) %>% 
  arrange(desc(n))


all_current %>% 
  filter(detention_facility_code == facil_code) %>%  
  count(msc_crime_class) %>% 
  arrange(desc(n)) %>% 
  mutate(prop = n / sum(n))

all_current %>% 
  filter(detention_facility_code != facil_code) %>%  
  count(msc_crime_class) %>% 
  arrange(desc(n)) %>% 
  mutate(prop = n / sum(n))

mean(as.numeric(all_current[all_current$detention_facility_code == facil_code, 'stay_length_min'], units='days'))
mean(as.numeric(all_current[all_current$detention_facility_code != facil_code, 'stay_length_min'], units='days'))

median(as.numeric(all_current[all_current$detention_facility_code == facil_code, 'stay_length_min'], units='days'))
median(as.numeric(all_current[all_current$detention_facility_code != facil_code, 'stay_length_min'], units='days'))


max(as.numeric(all_current[all_current$detention_facility_code == facil_code, 'stay_length_min'], units='days'))
min(as.numeric(all_current[all_current$detention_facility_code == facil_code, 'stay_length_min'], units='days'))

p1 <- all_current %>% 
  mutate(facil = case_when(detention_facility_code == facil_code ~ facil_code,
                           detention_facility_code != facil_code ~ "ALL OTHERS"),
         stay_length_min = as.numeric(stay_length_min, unit = "days")) %>% 
  group_by(facil) %>% 
  ggplot(aes(x = facil, y = log(stay_length_min), color=facil)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0, hjust=0))
  
p1


p2 <- all_current %>% 
  mutate(facil = case_when(detention_facility_code == facil_code ~ facil_code,
                           detention_facility_code != facil_code ~ "ALL OTHERS"),
         citizenship_country = case_when(citizenship_country == "CHINA, PEOPLES REPUBLIC OF" ~ "CHINA",
                                         citizenship_country %in% unlist(top_cit$citizenship_country) ~ citizenship_country,
                                         TRUE ~ "ALL OTHERS"),
         stay_length_min = as.numeric(stay_length_min, unit = "days")) %>% 
  group_by(facil, citizenship_country) %>% 
  ggplot(aes(x = citizenship_country, y = log(stay_length_min), color=facil)) +
  geom_boxplot() + 
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
  

p2

p3 <- all_current %>% 
  mutate(facil = case_when(detention_facility_code == facil_code ~ facil_code,
                           detention_facility_code != facil_code ~ "ALL OTHERS"),
         stay_length_min = as.numeric(stay_length_min, unit = "days")) %>% 
  group_by(facil, msc_crime_class) %>% 
  ggplot(aes(y = msc_crime_class, x = log(stay_length_min), color=facil)) +
  geom_boxplot()
  

p3

p4 <- unique_stays %>% 
  filter(stay_length_min > 0) %>% 
  mutate(facil = case_when(detention_facility_code == facil_code ~ facil_code,
                           detention_facility_code != facil_code ~ "ALL OTHERS"),
         stay_length_min = as.numeric(stay_length_min, unit = "days")) %>% 
  group_by(facil, msc_crime_class) %>% 
  ggplot(aes(y = msc_crime_class, x = log(stay_length_min), color=facil)) +
  geom_boxplot()

p4


```

```{r final_program}

p1 <- facil_unique_stays %>% 
  filter(stay_book_in_date_time >= "2011-10-01") %>% 
  mutate(detention_fy = substr(quarter(stay_book_in_date_time, with_year = TRUE, fiscal_start = 10), 1,4))%>% 
  count(detention_fy, arresting_authority) %>%
  ggplot(aes(x=as.factor(detention_fy), y=n, fill=arresting_authority)) +
  geom_col()

p1

```

```{r arresting_auth_facil_vs_nat}

facil_dat <- facil_unique_stays %>% 
  filter(stay_book_in_date_time >= "2011-10-01") %>% 
  mutate(detention_fy = substr(quarter(stay_book_in_date_time, with_year = TRUE, fiscal_start = 10), 1,4))%>% 
  count(detention_fy, arresting_authority) %>% 
  mutate(group = facil_code)

natl_dat <- unique_stays %>% 
  filter(stay_book_in_date_time >= "2011-10-01") %>% 
  mutate(detention_fy = substr(quarter(stay_book_in_date_time, with_year = TRUE, fiscal_start = 10), 1,4))%>% 
  count(detention_fy, arresting_authority) %>% 
  mutate(group = "national")

dat <- rbind(facil_dat, natl_dat)

dat %>% 
  ggplot(aes(x = detention_fy, y = n, fill=arresting_authority)) +
  geom_col(position='fill') +
  facet_wrap(~group)

```

## Facil characteristic headcounts

```{r headcounter}

min_date <- min(df$stay_book_in_date_time, na.rm=TRUE)
max_date <- max(df$stay_book_out_date_time, na.rm=TRUE)

timeline <- seq(min_date, max_date, by='day')

# Additional argument allowing us to specify which dataframe to apply function to. 
# We also move some dataframe setup requirements into function
# `timeline` needs to exist outside of function
headcounter <- function(date, df=df, var=var) {
  df[[var]] <- factor(df[[var]], levels = sort(unique(df[[var]])))
  df <- df %>%
    mutate(detention_book_out_date_time_min = case_when(is.na(detention_book_out_date_time) ~ max_date,
                                                    TRUE ~ detention_book_out_date_time))
  df[df$detention_book_in_date_and_time <= date & df$detention_book_out_date_time_min >= date,] %>% 
  count(.data[[var]]) %>% 
  complete(.data[[var]], fill = list(n = 0)) %>% 
  arrange(.data[[var]]) %>% 
  mutate(date=date)
  }

```

```{r citizenship_headcount}

system.time({facil_cit_headcount <- lapply(timeline, headcounter, df=facil_records, var='citizenship_country')})

facil_cit_headcount_data <- map_dfr(facil_cit_headcount, bind_rows)

facil_top_citizenship <- facil_unique_stays %>% 
  count(citizenship_country) %>% 
  arrange(desc(n))

p1 <- facil_cit_headcount_data %>% 
  mutate(citizenship_country = case_when(citizenship_country %in% unlist(head(facil_top_citizenship, 10)) ~ citizenship_country,
                                         TRUE ~ "ALL OTHERS")) %>% 
  group_by(date, citizenship_country) %>% 
  summarize(total_detained = sum(n)) %>% 
  filter(date > "2011-10-01") %>% 
  ggplot(aes(x = date, y = total_detained, color=citizenship_country, fill=citizenship_country, group=citizenship_country)) +
  geom_line() +
  labs(title = "Daily pop. per citizenship country",
       subtitle = facil_name)

p1

p1.1 <- facil_cit_headcount_data %>% 
  mutate(citizenship_country = case_when(citizenship_country %in% unlist(head(facil_top_citizenship, 10)) ~ citizenship_country,
                                         TRUE ~ "ALL OTHERS")) %>% 
  group_by(date, citizenship_country) %>% 
  summarize(total_detained = sum(n)) %>% 
  mutate(year_mth = as.yearmon(date)) %>% 
  group_by(year_mth, citizenship_country) %>% 
  summarize(adp = mean(total_detained, na.rm=TRUE)) %>% 
  filter(year_mth > "2011-10-01") %>% 
  ggplot(aes(x = year_mth, y = adp, color=citizenship_country, fill=citizenship_country, group=citizenship_country)) +
  geom_col(position="fill") +
  labs(title = "Monthly ADP per citizenship country",
       subtitle = facil_name)

p1.1

p1.2 <- facil_cit_headcount_data %>% 
  mutate(citizenship_country = case_when(citizenship_country %in% unlist(head(facil_top_citizenship, 10)) ~ citizenship_country,
                                         TRUE ~ "ALL OTHERS")) %>% 
  group_by(date, citizenship_country) %>% 
  summarize(total_detained = sum(n)) %>% 
  mutate(fy = substr(quarter(date, with_year = TRUE, fiscal_start = 10), 1, 4)) %>% 
  group_by(fy, citizenship_country) %>% 
  summarize(adp = mean(total_detained, na.rm=TRUE)) %>% 
  filter(fy > "2011-10-01") %>% 
  ggplot(aes(x = fy, y = adp, color=citizenship_country, fill=citizenship_country, group=citizenship_country)) +
  geom_col() +
  labs(title = "Annual ADP per citizenship country",
       subtitle = facil_name)

p1.2

facil_top_citizenship <- facil_unique_stays %>% 
  filter(detention_book_in_date_and_time > "2019-10-01") %>% 
  count(citizenship_country) %>% 
  arrange(desc(n))


p2 <- facil_cit_headcount_data %>% 
  mutate(citizenship_country = case_when(citizenship_country %in% unlist(head(facil_top_citizenship, 10)) ~ citizenship_country,
                                         TRUE ~ "ALL OTHERS")) %>% 
  group_by(date, citizenship_country) %>% 
  summarize(total_detained = sum(n)) %>% 
  filter(date > "2019-10-01") %>% 
  ggplot(aes(x = date, y = total_detained, color=citizenship_country, fill=citizenship_country, group=citizenship_country)) +
  geom_line() +
  labs(title = "Daily pop. per citizenship country",
       subtitle = facil_name)

p2

p2.1 <- facil_cit_headcount_data %>% 
  mutate(citizenship_country = case_when(citizenship_country %in% unlist(head(facil_top_citizenship, 10)) ~ citizenship_country,
                                         TRUE ~ "ALL OTHERS")) %>% 
  group_by(date, citizenship_country) %>% 
  summarize(total_detained = sum(n)) %>% 
  mutate(year_mth = as.yearmon(date)) %>% 
  group_by(year_mth, citizenship_country) %>% 
  summarize(adp = mean(total_detained, na.rm=TRUE)) %>% 
  filter(year_mth > "2019-10-01") %>% 
  ggplot(aes(x = year_mth, y = adp, color=citizenship_country, fill=citizenship_country, group=citizenship_country)) +
  geom_line() +
  labs(title = "Monthly ADP per citizenship country",
       subtitle = facil_name)

p2.1

```

```{r final_program_headcount}

system.time({facil_fp_headcount <- lapply(timeline, headcounter, df=facil_records, var='final_program')})

facil_fp_headcount_data <- map_dfr(facil_fp_headcount, bind_rows)

facil_top_program <- facil_unique_stays %>% 
  count(final_program) %>% 
  arrange(desc(n))

p1 <- facil_fp_headcount_data %>% 
  mutate(final_program = case_when(final_program %in% unlist(head(facil_top_program, 5)) ~ final_program,
                                         TRUE ~ "ALL OTHERS")) %>% 
  group_by(date, final_program) %>% 
  summarize(total_detained = sum(n)) %>% 
  filter(date > "2011-10-01") %>% 
  ggplot(aes(x = date, y = total_detained, color=final_program, fill=final_program, group=final_program)) +
  geom_col() + 
  labs(title = "Daily population per `final_program`",
       subtitle = facil_name)

p1

```

```{r headcount_experiment}

system.time({facil_exp_headcount <- lapply(timeline, headcounter, df=facil_records, var='first_facil')})

facil_exp_headcount_data <- map_dfr(facil_exp_headcount, bind_rows)

facil_top_facil <- facil_unique_stays %>% 
  count(first_facil) %>% 
  arrange(desc(n))

p1 <- facil_exp_headcount_data %>% 
  group_by(date, first_facil) %>% 
  summarize(total_detained = sum(n)) %>% 
  filter(date > "2011-10-01") %>% 
  ggplot(aes(x = date, y = total_detained, color=first_facil, fill=first_facil, group=first_facil)) +
  geom_line()

ggplotly(p1)


active_facil <- facil_exp_headcount_data %>% 
  mutate(year_mth = as.yearmon(date)) %>% 
  filter(n > 0) %>% 
  group_by(year_mth) %>% 
  summarize(n_facil = n_distinct(first_facil))

p2 <- active_facil %>% 
  filter(year_mth >= "2011-10-01") %>% 
  ggplot(aes(x = year_mth, y = n_facil)) +
  geom_col()

p2

```

```{r entry_date}

facil_unique_stays$entry_date <- as.Date(facil_unique_stays$entry_date, format="%Y-%m-%d")

p1 <- facil_unique_stays %>% 
  filter(entry_date < as.Date("2024-12-31"),
         entry_date > as.Date("1945-01-01"),
         detention_book_in_date_and_time > as.Date("2011-10-01")) %>% 
  mutate(detention_fy = substr(quarter(detention_book_in_date_and_time, fiscal_start = 10, type="year.quarter"), 1, 4)) %>% 
  ggplot(aes(x = detention_fy, y = entry_date)) +
  geom_boxplot()

p1

facil_unique_stays$diff_entry_det <- facil_unique_stays$entry_date - 
  as.Date(paste(year(facil_unique_stays$detention_book_in_date_and_time),
                 month(facil_unique_stays$detention_book_in_date_and_time),
                 day(facil_unique_stays$detention_book_in_date_and_time),
                 sep='-'))

p2 <- facil_unique_stays %>% 
  filter(entry_date < as.Date("2024-12-31"),
         entry_date > as.Date("1945-01-01"),
         detention_book_in_date_and_time > as.Date("2011-10-01")) %>% 
  mutate(detention_fy = substr(quarter(detention_book_in_date_and_time, fiscal_start = 10, type="year.quarter"), 1, 4)) %>% 
  ggplot(aes(x = detention_fy, y = abs(as.numeric(diff_entry_det)))) +
  scale_y_log10() +
  geom_boxplot()

p2

p3 <- facil_unique_stays %>% 
  filter(entry_date < as.Date("2024-12-31"),
         entry_date > as.Date("1945-01-01"),
         detention_book_in_date_and_time > as.Date("2011-10-01")) %>% 
  ggplot(aes(x = detention_book_in_date_and_time, y = entry_date)) +
  geom_point()

p3


```

