---
title: "FY12-24YTD ICE detention facility network"
author: "UWCHR"
date: "2025-02-19"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

One use case for ICE detention history data is to analyze networks of detention transfers. In the below we set up a very basic framework for such analysis.

```{r import_data, message=FALSE, warning=FALSE, include=TRUE}
options(scipen = 1000000)

library(pacman)
p_load(tidyverse, arrow, lubridate, zoo, digest, ggplot2, plotly, gghighlight, knitr, kableExtra, igraph, ggraph)

df <- data.table::fread(file = here::here('export', 'output', 'ice_detentions_fy12-24ytd.csv.gz'),
                        sep = '|')

data.table::setDF(df)

detloc_aor <- df %>% 
  distinct(detention_facility_code, area_of_responsibility) %>% 
  arrange(detention_facility_code, area_of_responsibility)


```

Generate a detention transfer "edge list" by lagging detention placements per `stayid`. Each row represents a facility transfer where `from_facil` is the previous placement `detention_facility_code` and `to_facil` is current placement `detention_facility_code`. Initial detention book-ins will have missing `from_facil` values. We can assume that placement start dates in `detention_book_in_date_and_time` represent the approximate date of transfer between facilities. Other placement characteristics can included in the `dplyr::select()` call or joined back to the resulting dataframe by `recid`.

```{r network_setup, message=FALSE, warning=FALSE, include=TRUE}

# Set up transfer "edge list" with one row per facility transfer
# Initial placements will have `from_facil` == NA
# Take placement start date `detention_book_in_date_and_time` as approximate date of transfer 
# Can bring back in other placement characteristics from full dataframe by joining with `recid`
transfers <- df %>% 
  dplyr::select(stayid, recid, stay_book_in_date_time, detention_book_in_date_and_time, detention_facility_code) %>% 
  group_by(stayid) %>% 
  mutate(from_facil = dplyr::lag(detention_facility_code, n = 1, default = NA),
         transfer_fy = as.numeric(substr(quarter(detention_book_in_date_and_time, with_year = TRUE, fiscal_start = 10), 1, 4))) %>% 
  rename(to_facil = detention_facility_code)

```

Very basic example implementation of network graph plot for nationwide facility transfers since FY2023; not immediately useful in this form, except to note that a small subsection is isolated from the broader network.

Numeric summary of transfers may be more informative than visual plot for many purposes. Network graph could be configured to represent real-world facility location by geocoding and joining in facility addresses from `analyze/input/facilities.csv.gz`.

Extensive documentation of network graph techniques: https://kateto.net/netscix2016.html

```{r network_graph, message=FALSE, warning=FALSE, include=TRUE}

# Summarize transfers since FY 2023
# Can also filter for all transfers involving specific facility, etc.
dat <- transfers %>% 
  filter(transfer_fy >= 2023,
         !is.na(from_facil),
         # to_facil == "CSCNWWA" | from_facil == "CSCNWWA"
         ) %>% 
  group_by(from_facil) %>% 
  count(to_facil)

# create the network object
network <- graph_from_data_frame(d=dat, directed=T)

# E(network)$width <- 1+E(network)$n/12

l <- layout_with_fr(network)

V(network)$size <- 8

V(network)$frame.color <- "white"

V(network)$color <- "orange"

V(network)$label <- "" 

E(network)$arrow.mode <- 0

# plot it
plot(network, layout=l)

```

